---
title: Visualizing Data in R with ggplot2 (Part 1)
author: Joseph Walker
date: '2018-11-26'
slug: visualizing-data-in-r-with-ggplot2-part-1
categories:
  - R
  - visualization
tags:
  - ggplot2
  - visualization
  - tutorials
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)

```

## This is the first of a three part series on data visualization using the popular `ggplot2` package.

**Check out [part 2](https://databeta.org/r/visualization/visualizing-data-in-r-with-ggplot2-part-2) or [part 3](https://databeta.org/r/visualization/visualizing-data-in-r-with-ggplot2-part-1) for more!**

# Part 1

Data Visualization combines statistics and design to present data in meaningful ways. Good design aids in both the understanding and communication of results. 

**Exploratory Visualizations** are meant to confirm and analyze. They are often data heavy, easy to generate, and intended for a small, specialist audience.

**Explanatory Visualizations** are meant to inform and persuade. They are labor intensive, data specific, and intended for a broader audience such as publications or presentations.

## The Grammar of Graphics

ggplot2's functionality relies on the grammar of graphics which has 2 principles: 
 
* Graphics are distinct layers of grammatical elements 
* Meaningful plots through aesthetic mapping 

The essential grammatical elements are: 
 
* Data - The dataset being plotted
* Aesthetics - The scales onto which we map our data
* Geometries - The visual elements used for our data

Other elements include: 

* Facets 
* Statistics 
* Coordinates 
* Themes 

## Aesthetics 

With ggplot2, plots become objects which can be recycled and manipulated by adding on layers and arguments. 
 
Aesthetics are mapped onto the plot using existing data. 
 
Common aesthetics include: 
 
* x 
* y 
* color 
* fill 
* size 
* alpha 
* linetype 
* labels 
* shape

```{r}
#subset/sample the diamonds dataset
diamonds_sample <- sample_n(diamonds, size = 100)

#Create a ggplot object price vs. carat for the Diamonds Dataset
p <- ggplot(data = diamonds_sample, mapping = aes(x = carat, y = price))

#plot the object
p
```

Notice that the ggplot object is plotted, but there is no geometry applied to the object so we don't see any of the data.

Here we will add one of the most common geometries, `geom_point` to the object to get a scatterplot.
```{r}
p + geom_point()

#Here we apply color as an aesthetic.
p + geom_point(aes(color = clarity))

#We can also add the aesthetic to the orignal ggplot object. It is best practice to keep the aesthetics in the same layer as much as possible. This will allow your plotting code to be clearer and more readable.
ggplot(data = diamonds_sample, mapping = aes(x = carat, y = price, color = clarity))+
  geom_point()
```

As noted above, there are a variety of different aesthetics we can apply to our plotting object. Be careful as the more aesthetics you add, the more complex your plot becomes and this is not necessarily a good thing.

```{r}
#plot diamonds_sample using multiple aesthetics
ggplot(diamonds_sample, aes(carat, price, color = clarity, shape = cut, size = color))+
  geom_point()
```

### Attributes
The aesthetic layers can be applied as **attributes**. It is important to know the difference between the two. Where aesthetics are mapped to specific data points on the plot based on the variables, attributes are applied to the entire plot object. 

```{r}
p <- ggplot(diamonds_sample, aes(carat, price, color = clarity))

p + geom_point(alpha = 0.6, shape = 18, size = 3)
```

Attributes override aesthetic layers as shown in the example below.
```{r}
#the color attribute over rides the color = clarity aesthetic
p + geom_point(alpha = 0.6, shape = 18, size = 3, color = "red")
```

### Overplotting

Often is this case that you will have to deal with overplotting:
 
* Large datasets 
* Imprecise data and so points are not clearly separated on your plot 
* Interval data (i.e. data appears at fixed values) 
* Aligned data values on a single axis. 

As we saw earlier, we can adjust the size, shape, and alpha layers to account for this:
```{r}
p <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl)))

#notice we have some overlapping points. We can do better.
p + geom_point(size = 4)

#adjust the shape
p + geom_point(size = 4, shape = 1)
```

Another solution to overplotting is the `geom_jitter` function. 

```{r}
p <- ggplot(mtcars, aes(x = cyl, y = wt))

p + geom_point()

# with geom_jitter
p + geom_jitter()
```

As shown above, `geom_jitter` fixed the overplotting, but it overcorrected. We can fix this using a variation on the function, calling the `width` argument.

```{r}
#set the width within the geom_jitter function
p + geom_jitter(width = 0.1)
```

And finally, we can pass the `position_jitter()` function to the `position = ` argument within `geom_point()`

```{r}
p + geom_point(position = position_jitter(0.1))
```

## Geometries

In the previous examples, we already saw one of the most common types of geometries used for plotting- points used for scatterplots - which use the `geom_point()` or `geom_jitter()` functions or arguments. 

Bars are another common geometry when it comes to data visualization.

**Histograms** are useful plots for visualizing distributions of a dataset and ggplot2 comes prepared with the `geom_histgoram` function. 

The default argument is `stat = 'bin'`, separates the continuous variable into bins so you get a sense of the general distribution of the data. The amount of bins defaults to `binwidth = range/30`.

```{r}
#plot histogram of the weight variable from the ChickWeight dataset
p <- ggplot(ChickWeight, aes(weight))
  
p + geom_histogram()

#adjust the binwidth
p + geom_histogram(binwidth = 5)
```

We can also customize the y-axis. By default, the y-axis displays the count of the dataset, but we can change it to display the density. Displaying the density is useful for showing the proprotional frequency of a bin relative to the whole dataset.

```{r}
p + geom_histogram(aes(y = ..density..), binwidth = 5, fill = "steelblue")
```

It could be the case that you want to split your distributions by some factor. To avoid overlap, `geom_histogram` stacks bars at each bin to display the distribution.

```{r}
#plot the chick weight distribution by diet
p <- ggplot(ChickWeight, aes(x = weight, fill = Diet))

p + geom_histogram(binwidth = 5)
```

As you can see, this can be a little difficult to discern. The y-axis count values are a sum of the distribution at that particular bin which can be misleading.  To correct this, we can set `position = "identity"`, which unstacks the bins.

```{r}
p + geom_histogram(binwidth = 5, position = "identity", alpha = 0.6)
```

An even better solution is to use `geom_freqpoly`.

```{r}
#remember to change fill to color 
ggplot(ChickWeight, aes(weight, color = Diet))+
  geom_freqpoly(binwidth = 5)
```

Another use of bars is the `geom_bar` used for bar plots. It has a `position` argument that can take on three arguments:
 
* stack (the default position)
* fill
* dodge

```{r}
#create barplot with the mtcars dataset
p <- ggplot(mtcars, aes(x = factor(cyl), fill = factor(am)))

#bar plot with default position = "stack"
p+geom_bar()

#position fill
p + geom_bar(position = "fill")

#position dodge
p + geom_bar(position = 'dodge')

#overlapping bars
p + geom_bar(position = position_dodge(width = 0.3), alpha = 0.6)
```

It is possible to use both x and y as aesthetics with `geom_bar()` by using the `stat = 'identity` argument.

Notice the error in the first example.
```{r, error=TRUE}
#create new variable avg_mpg which calculates the average mpg by cyl and am. assign to ggplot object.
p <- mtcars %>%
  group_by(cyl, am) %>%
  summarise(avg_mpg = mean(mpg)) %>%
ggplot(aes(x = factor(cyl),y = avg_mpg, fill = factor(am)))

#error message due to y aesthetic 
p + geom_bar()

#use stat = 'identity' to create bar plot of the avg_mpg for each cyl, by am.
p + geom_bar(stat = 'identity', position = "dodge")
```

Finally, we'll have a look at line plots using the `geom_line` function. 

Line plots are useful for time-series analysis. We'll take a look at some examples using the `economics` dataset included in the `ggplot2` package.

```{r}
#have a look at the economics dataset
str(economics)

p <- ggplot(economics, aes(x = date, y = unemploy))

p + geom_line()

#create df recess of recession dates    
recess <- data.frame(begin = as.Date(c('1969-12-01', '1973-11-01', '1980-01-01', '1981-07-01', '1990-07-01', '2001-03-01')),
                     end = as.Date(c('1970-11-01', '1975-03-01', '1980-07-01', '1982-11-01', '1991-03-01', '2001-11-01')))


#plot the periods of recession. 
#Note the inherit.aes = FALSE argument
ggplot(economics, aes(x = date, y = unemploy)) + 
  geom_line() +
  geom_rect(data = recess, aes(xmin = begin, xmax = end, ymin = -Inf, ymax = Inf), fill = "red", alpha = 0.2, inherit.aes = FALSE)
  

```

And just like other geometries, `geom_line` can take on various aesthetics/attributes. Let's explore these with the `txhousing` dataset from the `ggplot2` package.

`geom_path` can also be used in substitute of `geom_line`

```{r}
#view the dataset
str(txhousing)

#manipulate the dataset
txhousing$city <-factor(txhousing$city)

#how many cities in the dataset?
levels(txhousing$city)

#subset the data for only a few cities
txh <- txhousing %>% filter(city %in% c("Austin", "El Paso", "Houston", "San Antonio", "Dallas", "Corpus Christi", "Texarkana", "Odessa", "Fort Worth", "Waco"))

#create summary statistics using txh dataset
txh_summary <- txh %>% 
  group_by(year, city) %>%
summarise(avg_sales_per_year = mean(sales))

#create plot object using txh_summary dataset
p <- ggplot(txh_summary, aes(x = year, y = avg_sales_per_year, color = city))

#line plot
p + geom_line()
  
#adjust the attributes of the line plot
p + geom_line(lineend = "round", alpha = 0.6, size = 2)

#more attributes!
p + geom_line(linetype = 5, arrow = arrow(angle = 20, ends = "last", type = "closed"))
```

```
