---
title: The Hedgehog Concept
author: JW
date: '2019-03-12'
slug: the-hedgehog-concept
categories: []
tags:
  - thoughts
  - hedgehog
  - venn diagram
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = FALSE, message = FALSE)
```
I recently listened to a podcast with Jim Collins, a somewhat reclusive polymath who is among many things, a brilliant thinker, entrepreneur, and writer on the topics of business and leadership. One of the ideas that really stuck with me is the "Hedgehog Concept" which is based on a famous essay by Isaiah Berlin, titled "The Hedgehog and the Fox".  The title is a reference to the ancient Greek poet Archilochus: “A fox knows many things, but the hedgehog knows one big thing”. Berlin's essay utilizes this metaphor to categorize various literary figures through the ages. On one side, there are the foxes who draw upon different experiences and ideas in which to view the world.  On the other, the hedgehogs are those who see the world through a single defining idea. For Berlin, Tolstoy is a bit of a mystery and Berlin spends a good portion of the essay analyzing him through his text War and Peace. 

From a business standpoint, Jim favors the hedgehog. In his book Good to Great, Jim uses the metaphor of the hedgehog to explain what propels businesses from being good to great. The Hedgehog Concept is not just a strategy. It is, in some regard, something much more simple. It is a fundamental understanding of the business itself and that ‘one big thing’ that lies at the intersection of three circles or concepts. 

The three circles are:

* What are you passionate about?
* What can you be the best in the world at?
* What drives your economic engine?

The Hedgehog Concept can also be applied personally to one’s own life. However, it may be wise to think about the circles in more practical terms (my thoughts, not Jim's).

```{r echo = F}
library(VennDiagram)

venn.plot <- draw.triple.venn(
  area1 = 65,
  area2 = 75,
  area3 = 85,
  n12 = 35,
  n23 = 15,
  n13 = 25,
  n123 = 5,
  category = c(paste("What am I \n interested in?"), paste("What am I \n good at?"), "What will people pay me to do?"),
  fill = c("salmon", "steelblue", "seagreen"),
  lty = "blank",
  cex = 0,
  cat.cex = 1.5,
  cat.default.pos = "text",
  cat.col = c("black", "black", "black") 
)

grid.draw(venn.plot)

grid.text(label = "The Sweet Spot")
```

The first circle. Following one's passion is not always the best advice and doesn't work for everyone. Rather, **what do you enjoy doing?** What gets you out of bed in the morning and makes the time fly by? This sort of thing is worth pursuing. 

This second circle is a tricky one. Being the best in the world at something is extremely rare and most of us never have the time or opportunity to discover that thing. But there are certainly things we are good at. Jim calls this the curse of competence. He argues that we can be good at lots of things simply because we have an abundance of talent. Yet that doesn't mean we were meant to do them. He firmly believes that there are things we were born to do, genetically encoded to be great, if not the best at. This is the focus of the second circle which we'll call: **what am I meant to do?** I strongly believe it's important not to get too fixated on this idea. It's important to try new things; to allow space and time to explore, discover, and most importantly, change. What we may think we're meant to do today may not hold a decade from now. And just because you were born to do it doesn't mean it still won't be hard at times. Everything worth doing is difficult at one time or another.

And finally, the third circle remains relatively consistent from a business standpoint. **What are people willing to pay you to do?** We all have bills to pay. Sure, I love climbing but I'm not the best climber out there and sponsors are not knocking down my door throwing free gear at me and sending me to remote destinations to climb. I wish. 

Presumably, the sweet spot lies at where these three circles overlap. Sounds about right to me. 

Here's some code in case you want to make your own venn diagram!

```{r eval = F, highlight=TRUE}
library(VennDiagram)

venn.plot <- draw.triple.venn(
  area1 = 65,
  area2 = 75,
  area3 = 85,
  n12 = 35,
  n23 = 15,
  n13 = 25,
  n123 = 5,
  category = c(paste("What am I \n interested in?"), paste("What am I \n good at?"), "What will people pay me to do?"),
  fill = c("salmon", "steelblue", "seagreen"),
  lty = "blank",
  cex = 0,
  cat.cex = 1.5,
  cat.default.pos = "text",
  cat.col = c("black", "black", "black") 
)

grid.draw(venn.plot)

grid.text(label = "The Sweet Spot")
grid.text(label = "The Hedgehog Concept", vjust = -20, gp = gpar(fontsize = 20))
```
