---
title: "An Introduction to `forcats`"
author: "Joseph Walker"
date: "March 22, 2017"
output: 
  html_document: 
    toc: yes
---

```{r global options, include = TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

###Defining Factors
The `forcats` package is part of the tidyverse and is useful for dealing with factors. Factors are simply categorical variables, useful for controlling the levels and order of a vector.

Categoricalv or discrete variables, as opposed to continuous variables, are often qualitiative and can take on a finite number of values.

Examples of categorical variables: types of fruit, locations, party preference, ethnicity.

```{r Example 1}
#colors is a character vector of length nine consisting of colors: red, blue, & green

colors <- c("red","blue","green","green","red","blue","red","green","blue")
class(colors)

#The factor() function works by assigning integers (number values) to the categorical values (red, blue, green) of the vector or variable. 

colors <- factor(colors)
print(colors)
class(colors)

str(colors)
#Each color is a level, and underlying each level is an integer associated with that level; red = 3, green = 2, blue = 1. The order of the levels is assigned alphabetically to the integer: b = 1, g = 2, r = 3.

#We can reorder the levels if we want to:
colors <- factor(colors, levels = c("red","blue", "green"))
print(colors)
#The order of the vector stays the same, but now the order of the levels has changed so that red = 1, blue = 2, green = 3
str(colors)
```

Factors are a useful data structure that allow you to have more control over how you analyze and visualize your data.
```{r Example 2}
#Let's compare the summary function for colors as a character vector and as a factor.
colors <- c("red","blue","green","green","red","blue","red","green","blue")

summary(colors)

colors <- factor(colors, levels = c("red","blue", "green"))

summary(colors)

#As factors, each color becomes a distinct group and summarized separately.
```

###The Forcats Package
As useful as factors are, they can be quite a pain to work with at times. Luckily, `forcats` has functions that allow you to manipulate factors. 


First, let's install the package using:
`install.packages("forcats")`

`forcats` also comes installed as part of the `tidyverse` suite of packages. It is not part of the core group of `tidyverse` packages so we must load it explicitly.

```{r install/load forcats}
library(tidyverse)
library(forcats)
```

###Function: Factor Recode
One useful function of `forcats` is `fct_recode`. This allows you to change the levels (or name/identity) of a factor. Here's an example:

```{r change level values}
#Let's use the airquality data set that comes pre-installed in R
glimpse(airquality)

#The month column is data type integer, let's change it to a factor 
airquality$Month <- factor(airquality$Month)

levels(airquality$Month)

#Let's rename the months using the fct_recode function. 
airquality$Month <- fct_recode(airquality$Month, May = '5', June = '6', July = '7', Aug = '8', Sept = '9')
glimpse(airquality$Month)

ggplot(airquality, aes(Month, Temp)) +
  geom_boxplot(aes(fill = Month)) +
  ggtitle(label = "Daily Temperatures Aggregated by Month")
```

###The Factor Reverse Function
If you just want to reverse the order, there's the `fct_rev` function. You can even use it in line when defining your aesthetics in ggplot like so:
```{r reverse the order}

ggplot(airquality, aes(fct_rev(Month), Temp)) +
  geom_boxplot(aes(fill = Month)) +
  labs(x = "Month") +
  ggtitle(label = "Our plot now has the x-axis in reverse order")
```

###The Factor Relevel Function
Another useful function is `fct_relevel`. This function allows us to change any number of levels to any position.
```{r change level position}
airquality$Month <- fct_relevel(airquality$Month, 'Sept', 'July', 'May', 'Aug', 'June')

levels(airquality$Month)

# This may not seem useful at first, but when you need to visualize your data in a particular way, the fct_relevel function is extremely useful...

ggplot(airquality, aes(Month, Temp)) +
  geom_boxplot(aes(fill = Month)) +
  ggtitle(label = "Notice how the order of the level 'Month' has changed")
```

###The Factor Reorder Function

And finally, it is often useful to reorder the factor in a way that is useful for visualization. For this, we can use the `fct_reorder` function.

For this example, let's use the mtcars data set:
```{r fct_reorder setup}
mtcars$model <- row.names(mtcars)

glimpse(mtcars)

mtcars$model <- factor(mtcars$model)

ggplot(mtcars, aes(mpg, model)) +
  geom_point() +
  ggtitle(label = "MPG vs. Car Model")

```

It's difficult to make comparisons when the data is scattered. But we're in luck! We can use the `fct_reorder` function to clean it up.

```{r fct_reorder}
#fct_reorder takes three arguments: f = factor you want to reorder, x = the variable in which the order will be based upon, and optionally fun (a function to  be used if there are multiple values of x for each value of f.) Here we focus on only the first two arguments.

ggplot(mtcars, aes(mpg, fct_reorder(f = model, x = mpg))) +
  geom_point() +
  labs(y = "model") +
  ggtitle(label = "We can make better comparison by reordering the levels based on the mpg values!") +
  theme(plot.title = element_text(size = 10, face = 'bold'))

```



